ProteInfer, deep neural networks for protein functional inference

### 2023-01-16 latex/draft3.tex
upd

\begin{figure}[h!]
  \centering
  \includegraphics[width=.4\columnwidth]{Figures/qr.pdf}

@@

\section*{Introduction}
Every day, more than a hundred thousand protein sequences are added to global sequence databases \cite{tremblstats}. However, these entries are of limited use to practitioners unless they are accompanied by functional annotations. While curators diligently extract annotations from the literature, assessing more than 60,000 papers each year \cite{uniprotlatest}, the time-consuming nature of this task means that only 0.03\% of publicly available protein sequences are manually annotated.
%%%FIGURETREMBL%%%
The community has a long history of using computational tools to infer protein function directly from amino acid sequence. Starting in the 1980s, methods such as BLAST \cite{blast} relied on pairwise sequence comparisons, where a query protein is assumed to have the same function as highly similar sequences that have already been annotated. Signature-based approaches were later introduced, with the PROSITE database \cite{bairoch1991prosite} cataloguing short amino acid ``motifs'' found in proteins that share a particular function. Subsequently, a crucial refinement of signature-based approaches was the development of profile hidden Markov models (HMMs) \cite{krogh1994hidden, eddyhmmreview}. These models collapse an alignment of related protein sequences into a model that provides likelihood scores for new sequences, that describe how well they fit the aligned set. Critically, profile HMMs allow for longer signatures and fuzzier matching and are currently used to update popular databases such as Interpro and Pfam \cite{pfamlatest, blum2021interpro}. Subsequent refinements have made these techniques more sensitive and computationally efficient \cite{altschul1997gapped, fasterhmmer, pfamlatest, soding2005protein, steinegger2019hh}, while their availability as web tools allows practitioners to easily incorporate them into workflows \cite{johnson2008ncbi, soding2005hhpred, potter2018hmmer, bernhofer2021predictprotein}.
\begin{figure}[t]
\centering
  \includegraphics[width=\columnwidth]{Figures/new_the_task.pdf}%
  \caption{\textbf{Three approaches for mapping from an amino acid sequence to inferred function:}   (1) finding similar sequences in a large database of sequences with known annotation (e.g., BLAST), (2) scoring against a large database of statistical models for each family of sequences with known function (e.g. InterProScan), and (3) applying a single deep neural network trained to predict multiple output categories (e.g., this work).\label{fig:task}}
\end{figure}
These computational modelling approaches have had great impact; however, one third of bacterial proteins still cannot be annotated (even computationally) with a function \cite{phenotypes}. It is therefore worthwhile to examine how new approaches might complement existing techniques. First, current approaches conduct entirely separate comparisons for each comparator sequence or model, and thus may not fully exploit the features shared across different functional classes. An ideal classification system, for example, might have a modular ATP-binding region detector used in detection of both kinases and ABC transporters \citep{pmid12468712}. Separately modelling these targets in each family, like standard HMM approaches, increases the computational cost and may also be less accurate. In addition, the process of creating many of these signatures is not fully automated and requires considerable curatorial efforts \citep{pfamprofilebuilding, pfamrepeats}, which at present are spread across an array of disparate but overlapping signature databases \citep{blum2021interpro}.

@@

\begin{figure}[h!]
\centering
  \includegraphics[width=\columnwidth]{Figures/figure.pdf}%
  \caption{\textbf{A deep dilated convolutional architecture for protein function prediction.} Amino acids are one-hot encoded, then pass through a series of convolutions implemented within residual blocks. Successive filters are increasingly dilated, allowing the top residual layer of the network to build up a representation of high-order protein features. The positional embeddings in this layer are collapsed by mean-pooling to a single embedding of the entire sequence, which is converted into probabilities of each functional classification through a fully connected layer with sigmoidal activations. }
  \label{architecture}
\end{figure}
\subsection*{A neural-network for protein function prediction}
%%%MODEL%%%
In a ProteInfer neural network (\Cref{architecture}%htmlver:<a href="#fig3" target="_self">Figure 3</a>
), a raw amino acid sequence is first represented numerically as a \textit{one-hot} matrix and then passed through a series of convolutional layers. Each layer takes the representation of the sequence in the previous layer and applies a number of \textit{filters}, which detect patterns of features. We use \textit{residual} layers, in which the output of each layer is added to its input to ease the training of deeper networks \cite{resnet}, and dilated convolutions \cite{dilatedconvolutions}, meaning that successive layers examine larger sub-sequences of the input sequence. After building up an embedding of each position in the sequence, the model collapses these down to a single $n$-dimensional embedding of the sequence using average pooling. Since natural protein sequences can vary in length by at least three orders of magnitude, this pooling is advantageous because it allows our model to accommodate sequences of arbitrary length without imposing restrictive modeling assumptions or computational burdens that scale with sequence length. In contrast, many previous approaches operate on fixed sequence lengths: these techniques are unable to make predictions for proteins larger than this sequence length, and use unnecessary resources when employed on smaller proteins. Finally, a fully-connected layer maps these embeddings to logits for each potential label, which are the input to an element-wise sigmoid layer that outputs per-label probabilities. 
We select all labels with predicted probability above a given confidence threshold, and varying this threshold yields a tradeoff between precision and recall. To summarize model performance as a single scalar, we compute the $\texttt{F}_{max}$ score, the maximum $\texttt{F}_{1}$ score (the geometric mean of precision and recall) across all thresholds \citep{radivojac2013large}.

@@

\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/fig3}%
  \caption{\textbf{ProteInfer performance (A) for all 7 top-level enzyme groups from a single CNN model (B) compared between methods: a single ProteInfer CNN, an ensemble of Proteinfer CNNs, a BLAST-based baseline, and an ensemble of BLAST predictions combined with ProteInfer CNNs }}
  \label{fig:toplevel}
\end{figure}

\begin{figure*}[t]
  \centering
    \includegraphics[width=\textwidth]{Figures/tetrahydrofolate_synthase.png}
    % See https://docs.google.com/presentation/d/1PqP-Ux7far9NMYvroeZIQQ9iXi41xira7-awDC4w_ss/edit#slide=id.p for the provenance of this figure
    \caption{\textbf{Linking sequence regions to function with class activation mapping for C-1-tetrahydrofolate synthase (accession P11586).} A. Ground truth annotation of function on Uniprot \citep{uniprotlatest}. B. The three horizontal bars are the sequence region ProteInfer predicts are most involved in each corresponding reaction. This concurs with the known function localization. }
    \label{fig:cam}
  \end{figure*}

  \begin{figure*}
    \centering
      \includegraphics[width=\textwidth]{Figures/umap3.pdf}%
      \caption{\textbf{Embedding reflects enzyme functional hierarchy}. UMAP projection of embeddings for the subset of test set sequences which have only one leaf-level EC classification. Points are colour-coded at successive levels of the EC hierarchy in each panel. (A) colours denote top level EC groups, (B) colours denote second level EC groups within EC2.*, (C) colours denote third level EC groups within EC:2.7.*, (D) colours depict terminal EC groups within EC:2.7.4.* }
      \label{fig:umap}
    \end{figure*}

@@

%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/nucleotidebinding}
        \caption{Nucleotide binding}
        \label{fig:gull}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/ribosome}
        \caption{Structural constituent of ribosome}
        \label{fig:tiger}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/membrane}
        \caption{Intrinsic component of membrane}
        \label{fig:mouse}
    \end{subfigure}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} This figure shows EC-trained ProteInfer embeddings for all non-enzymatic sequences in the test set, projected using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, \emph{Q8RUD6} is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). \emph{Q8RUD6} can however act as a sulfertransferase \cite{pmid17408957}.

@@

% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}[t]
\centering
  \includegraphics[width=\columnwidth]{Figures/price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}

@@

\clearpage
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/ec_label_distribution.png}
  \caption{Histogram of number of labels per sequence, including hierarchical labels, on the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/go_label_distribution.png}
  \caption{Histogram of number of labels per sequence, including hierarchical labels, on the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_ec_random.png}
  \caption{Number of sequences annotated with a given functional label. (EC class) in the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_go_random.png}
  \caption{Number of sequences annotated with a given functional label. (GO label) in the random dataset.}per p
\end{figure}
\pagebreak
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_ec_clustered.png}
  \caption{Number of sequences annotated with a given functional label. (EC class) in the clustered dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_go_clustered.png}
  \caption{Number of sequences annotated with a given functional label. (GO label) in the clustered dataset.}
\end{figure}

@@

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/random_truncated.pdf}
    \caption{EC number prediction: random split} 
    \label{fig:with_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/random_truncated_go.pdf}
    \caption{GO term prediction: random split} 
    \label{fig:with_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/clustered_truncated.pdf}
    \caption{EC number prediction: clustered split} 
    \label{fig:with_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/clustered_truncated_go.pdf}
    \caption{GO term prediction: clustered split} 
    \label{fig:with_ensemble:d} 
  \end{subfigure} 
  \caption{Bootstrapped precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs, and ensembled ProteInfer CNNs scaled by BLAST score. }
  \label{fig:with_ensemble} 
\end{figure}

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_ec_random_unzoomed.png}
    \caption{EC number prediction: random split} 
    \label{fig:without_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_go_random_unzoomed.png}
    \caption{GO term prediction: random split} 
    \label{fig:without_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_ec_clustered_unzoomed.png}
    \caption{EC number prediction: clustered split} 
    \label{fig:without_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_go_clustered_unzoomed.png}
    \caption{GO term prediction: clustered split} 
    \label{fig:without_ensemble:d} 
  \end{subfigure} 
  \caption{Full precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs }
  \label{fig:without_ensemble} 
\end{figure}

\begin{figure}[htbp]
\centering
  \includegraphics[width=0.6\textwidth]{Figures/random_untruncated.pdf}
  \caption{EC random task with different methods compared against a naive baseline where the predictor is simply the frequency in the training set.}
  \label{sup:fig:naive}
\end{figure}

@@

\subsection{GO performance by ontology}
\begin{figure}[htbp]
\centering
  \includegraphics[width=0.7\textwidth]{Figures/combined_split_go_random.pdf}
  \caption{GO performance stratified by method and ontology type.}
  \label{sup:fig:go_ontologies}
\end{figure}

\clearpage
\subsection{Stratified performance}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/ec_f1_by_training_size.png}
  \caption{Performance of EC model stratified by number of training examples available for each test example.}
  \label{sup:fig:ec_f1_by_training_size}
\end{figure}

@@

\begin{figure}[htbp]
\centering
  \includegraphics[width=0.7\textwidth]{Figures/pfam_random.png}
  \caption{Performance for Pfam prediction for random train/test split.}
  \label{sup:fig:ec_f1_by_training_size}
\end{figure}

>>>>

\begin{figure}[h!]
  \centering
  \includegraphics[width=.4\columnwidth]{qr.pdf}

@@

\section*{Introduction}
Every day, more than a hundred thousand protein sequences are added to global sequence databases \cite{tremblstats}. However, these entries are of limited use to practitioners unless they are accompanied by functional annotations. While curators diligently extract annotations from the literature, assessing more than 60,000 papers each year \cite{uniprotlatest}, the time-consuming nature of this task means that only 0.03\% of publicly available protein sequences are manually annotated.
%%%FIGURETREMBL%%%
The community has a long history of using computational tools to infer protein function directly from amino acid sequence. Starting in the 1980s, methods such as BLAST \cite{blast} relied on pairwise sequence comparisons, where a query protein is assumed to have the same function as highly similar sequences that have already been annotated. Signature-based approaches were later introduced, with the PROSITE database \cite{bairoch1991prosite} cataloguing short amino acid ``motifs'' found in proteins that share a particular function. Subsequently, a crucial refinement of signature-based approaches was the development of profile hidden Markov models (HMMs) \cite{krogh1994hidden, eddyhmmreview}. These models collapse an alignment of related protein sequences into a model that provides likelihood scores for new sequences, that describe how well they fit the aligned set. Critically, profile HMMs allow for longer signatures and fuzzier matching and are currently used to update popular databases such as Interpro and Pfam \cite{pfamlatest, blum2021interpro}. Subsequent refinements have made these techniques more sensitive and computationally efficient \cite{altschul1997gapped, fasterhmmer, pfamlatest, soding2005protein, steinegger2019hh}, while their availability as web tools allows practitioners to easily incorporate them into workflows \cite{johnson2008ncbi, soding2005hhpred, potter2018hmmer, bernhofer2021predictprotein}.
\begin{figure}[t]
\centering
  \includegraphics[width=\columnwidth]{new_the_task.pdf}%
  \caption{\textbf{Three approaches for mapping from an amino acid sequence to inferred function:}   (1) finding similar sequences in a large database of sequences with known annotation (e.g., BLAST), (2) scoring against a large database of statistical models for each family of sequences with known function (e.g. InterProScan), and (3) applying a single deep neural network trained to predict multiple output categories (e.g., this work).\label{fig:task}}
\end{figure}
These computational modelling approaches have had great impact; however, one third of bacterial proteins still cannot be annotated (even computationally) with a function \cite{phenotypes}. It is therefore worthwhile to examine how new approaches might complement existing techniques. First, current approaches conduct entirely separate comparisons for each comparator sequence or model, and thus may not fully exploit the features shared across different functional classes. An ideal classification system, for example, might have a modular ATP-binding region detector used in detection of both kinases and ABC transporters \citep{pmid12468712}. Separately modelling these targets in each family, like standard HMM approaches, increases the computational cost and may also be less accurate. In addition, the process of creating many of these signatures is not fully automated and requires considerable curatorial efforts \citep{pfamprofilebuilding, pfamrepeats}, which at present are spread across an array of disparate but overlapping signature databases \citep{blum2021interpro}.

@@

\begin{figure}[h!]
\centering
  \includegraphics[width=\columnwidth]{figure.pdf}%
  \caption{\textbf{A deep dilated convolutional architecture for protein function prediction.} Amino acids are one-hot encoded, then pass through a series of convolutions implemented within residual blocks. Successive filters are increasingly dilated, allowing the top residual layer of the network to build up a representation of high-order protein features. The positional embeddings in this layer are collapsed by mean-pooling to a single embedding of the entire sequence, which is converted into probabilities of each functional classification through a fully connected layer with sigmoidal activations. }
  \label{architecture}
\end{figure}
\subsection*{A neural-network for protein function prediction}
%%%MODEL%%%
In a ProteInfer neural network (\Cref{architecture}%htmlver:<a href="#fig3" target="_self">Figure 3</a>
), a raw amino acid sequence is first represented numerically as a \textit{one-hot} matrix and then passed through a series of convolutional layers. Each layer takes the representation of the sequence in the previous layer and applies a number of \textit{filters}, which detect patterns of features. We use \textit{residual} layers, in which the output of each layer is added to its input to ease the training of deeper networks \cite{resnet}, and dilated convolutions \cite{dilatedconvolutions}, meaning that successive layers examine larger sub-sequences of the input sequence. After building up an embedding of each position in the sequence, the model collapses these down to a single $n$-dimensional embedding of the sequence using average pooling. Since natural protein sequences can vary in length by at least three orders of magnitude, this pooling is advantageous because it allows our model to accommodate sequences of arbitrary length without imposing restrictive modeling assumptions or computational burdens that scale with sequence length. In contrast, many previous approaches operate on fixed sequence lengths: these techniques are unable to make predictions for proteins larger than this sequence length, and use unnecessary resources when employed on smaller proteins. Finally, a fully-connected layer maps these embeddings to logits for each potential label, which are the input to an element-wise sigmoid layer that outputs per-label probabilities. 
We select all labels with predicted probability above a given confidence threshold, and varying this threshold yields a tradeoff between precision and recall. To summarize model performance as a single scalar, we compute the $\texttt{F}_{max}$ score, the maximum $\texttt{F}_{1}$ score (the geometric mean of precision and recall) across all thresholds \citep{radivojac2013large}.

@@

\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{fig3}%
  \caption{\textbf{ProteInfer performance (A) for all 7 top-level enzyme groups from a single CNN model (B) compared between methods: a single ProteInfer CNN, an ensemble of Proteinfer CNNs, a BLAST-based baseline, and an ensemble of BLAST predictions combined with ProteInfer CNNs }}
  \label{fig:toplevel}
\end{figure}

\begin{figure*}[t]
  \centering
    \includegraphics[width=\textwidth]{tetrahydrofolate_synthase.png}
    % See https://docs.google.com/presentation/d/1PqP-Ux7far9NMYvroeZIQQ9iXi41xira7-awDC4w_ss/edit#slide=id.p for the provenance of this figure
    \caption{\textbf{Linking sequence regions to function with class activation mapping for C-1-tetrahydrofolate synthase (accession P11586).} A. Ground truth annotation of function on Uniprot \citep{uniprotlatest}. B. The three horizontal bars are the sequence region ProteInfer predicts are most involved in each corresponding reaction. This concurs with the known function localization. }
    \label{fig:cam}
  \end{figure*}

  \begin{figure*}
    \centering
      \includegraphics[width=\textwidth]{umap3.pdf}%
      \caption{\textbf{Embedding reflects enzyme functional hierarchy}. UMAP projection of embeddings for the subset of test set sequences which have only one leaf-level EC classification. Points are colour-coded at successive levels of the EC hierarchy in each panel. (A) colours denote top level EC groups, (B) colours denote second level EC groups within EC2.*, (C) colours denote third level EC groups within EC:2.7.*, (D) colours depict terminal EC groups within EC:2.7.4.* }
      \label{fig:umap}
    \end{figure*}

@@

%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{nucleotidebinding}
        \caption{Nucleotide binding}
        \label{fig:gull}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{ribosome}
        \caption{Structural constituent of ribosome}
        \label{fig:tiger}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{membrane}
        \caption{Intrinsic component of membrane}
        \label{fig:mouse}
    \end{subfigure}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} This figure shows EC-trained ProteInfer embeddings for all non-enzymatic sequences in the test set, projected using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, \emph{Q8RUD6} is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). \emph{Q8RUD6} can however act as a sulfertransferase \cite{pmid17408957}.

@@

% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}[t]
\centering
  \includegraphics[width=\columnwidth]{price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}

@@

\clearpage
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{ec_label_distribution.png}
  \caption{Histogram of number of labels per sequence, including hierarchical labels, on the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{go_label_distribution.png}
  \caption{Histogram of number of labels per sequence, including hierarchical labels, on the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{sequences_per_label_ec_random.png}
  \caption{Number of sequences annotated with a given functional label. (EC class) in the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{sequences_per_label_go_random.png}
  \caption{Number of sequences annotated with a given functional label. (GO label) in the random dataset.}per p
\end{figure}
\pagebreak
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{sequences_per_label_ec_clustered.png}
  \caption{Number of sequences annotated with a given functional label. (EC class) in the clustered dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{sequences_per_label_go_clustered.png}
  \caption{Number of sequences annotated with a given functional label. (GO label) in the clustered dataset.}
\end{figure}

@@

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{random_truncated.pdf}
    \caption{EC number prediction: random split} 
    \label{fig:with_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{random_truncated_go.pdf}
    \caption{GO term prediction: random split} 
    \label{fig:with_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{clustered_truncated.pdf}
    \caption{EC number prediction: clustered split} 
    \label{fig:with_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{clustered_truncated_go.pdf}
    \caption{GO term prediction: clustered split} 
    \label{fig:with_ensemble:d} 
  \end{subfigure} 
  \caption{Bootstrapped precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs, and ensembled ProteInfer CNNs scaled by BLAST score. }
  \label{fig:with_ensemble} 
\end{figure}

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{prc_ec_random_unzoomed.png}
    \caption{EC number prediction: random split} 
    \label{fig:without_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{prc_go_random_unzoomed.png}
    \caption{GO term prediction: random split} 
    \label{fig:without_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{prc_ec_clustered_unzoomed.png}
    \caption{EC number prediction: clustered split} 
    \label{fig:without_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{prc_go_clustered_unzoomed.png}
    \caption{GO term prediction: clustered split} 
    \label{fig:without_ensemble:d} 
  \end{subfigure} 
  \caption{Full precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs }
  \label{fig:without_ensemble} 
\end{figure}

\begin{figure}[htbp]
\centering
  \includegraphics[width=0.6\textwidth]{random_untruncated.pdf}
  \caption{EC random task with different methods compared against a naive baseline where the predictor is simply the frequency in the training set.}
  \label{sup:fig:naive}
\end{figure}

@@

\subsection{GO performance by ontology}
\begin{figure}[htbp]
\centering
  \includegraphics[width=0.7\textwidth]{combined_split_go_random.pdf}
  \caption{GO performance stratified by method and ontology type.}
  \label{sup:fig:go_ontologies}
\end{figure}

\clearpage
\subsection{Stratified performance}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{ec_f1_by_training_size.png}
  \caption{Performance of EC model stratified by number of training examples available for each test example.}
  \label{sup:fig:ec_f1_by_training_size}
\end{figure}

@@

\begin{figure}[htbp]
\centering
  \includegraphics[width=0.7\textwidth]{pfam_random.png}
  \caption{Performance for Pfam prediction for random train/test split.}
  \label{sup:fig:ec_f1_by_training_size}
\end{figure}
